Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences

### 2017-03-10 paper/AMIA/main.tex
final version by Alex

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Mind Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

\institutes{
$^1$Department of Computer Science,  
$^2$Pediatric Prevention Research Center, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics. In this paper, we focus on automating the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication exchanges to elicit a particular type of patient response and dynamically predicting this type of response in an observed patient-provider communication. Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov Chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity. It achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences using first and second order Markov Chain and Hidden Markov Model, respectively,  to model communication sequences. Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success. Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of effective behavioral interventions.}

@@

In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to designing an effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual annotation and examination of transcripts is a very time-consuming process, development of new obesity interventions can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions.

While our previous work \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing 115 patient and provider behavior codes \cite{carcone2013provider}, in the work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.

While sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to establish the relationship between physician and patient behaviors associated with patients' exhibiting cues, there have been only a handful of previous studies on modeling annotated sequences of patient-provider communication exchanges and predicting a desired patient behavior in the context of clinical interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances. Each motivational interview consists of counselor-adolescent and counselor-caregiver sessions. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 distinct behavior codes, which are grouped into the youth, caregiver, and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in either positive change talk or commitment language statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk, negative commitment language or the ones in which neither change talk nor commitment language statements occur. Out of 1072 observed sequences, 880 were positive and 192 were negative. alternating sequences are the possible combination of a natural sequence where two consecutive utterances cannot belong to the same speaker. For example, there are 6 possible alternating sequences ({C1, A1, C2, A3}, {C1, A1, C3, A3}, {C1, A1, C4, A3}, {C1, A2, C2, A3}, {C1, A2, C3, A3} and {C1, A2, C4, A3}) for the natural sequence {C1, A1, A2, C2, C3, C4, A3}, where adolescent and counselor codes are represented by A and C, respectively. A total of 18753 alternating sequences have been observed, out of which 18139 were successful and 614 were unsuccessful ones. For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences. Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
 natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
  \end{tabular}
\end{table}

@@

\subsection*{\textit{Sequence classification techniques}}

Generally, a sequence is a temporally ordered set of events. In our study, an event is a behavior code represented as a symbolic value, such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$, where $L$ is fixed and known in advance. In our case, the labels are ``successful'' and ``unsuccessful'' (motivational interview).

We model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are probabilistic methods for modeling discrete observation sequences with finite vocabulary. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.

\textbf {Markov Chain}: In probability theory, a Markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the Markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

The likelihood of success of a given motivational interview at a given time point, given a sequence of MI behavior codes observed prior to that point, is estimated based on the following formula:

\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}

if $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language. Estimating the model on the training data involves experimentally determining the value of $\delta$, resulting in the highest prediction F-measure with lower bias. Figure~\ref{fig:delta} illustrates the accuracy of Markov model-based method by varying the value of $\delta$. As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias. However, F-score sharply decreases after $\delta = -0.9$. Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.

@@

The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code to compute the state transition probability matrix. A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular state (symbol) $x_i$ is computed by looking at the k preceding states (symbols). Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols. In our experiment, we also used second order Markov model. Hence, our model has $N^2$ states with transition probability matrix of size $N^2 \times N$.

\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations. An HMM can be fully specified by the following quintuple:

\begin{center}
$\lambda = (N, M, A, B, \pi)$
\end{center}

@@

We exclude the structure parameters M and N to designate HMM using a compact notation:

\begin{center}
$\lambda = (A, B, \pi)$
\end{center}

The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}. As in the case of MC, we also experimented with second-order HMM.

\subsection*{\textit{Evaluation methods}}
The performance of classifiers is reported on the basis of 5-fold cross-validation by using weighted macro-averaging over folds. Precision, recall, F-measure and accuracy were used to evaluate the performance of classifiers. A true positive (TP) was counted when classifier correctly classified the sequence into actual class; a false positive (FP) was counted when classifier incorrectly classified other classes to the considered class; a false negative (FN) when the considered class incorrectly classified to other classes. The precision of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences identified as that particular class by the classifier. In other words, Precision = TP / (TP + FP). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (labeled data set). In more specifically, Recall = TP / (TP + FN). Similarly, Accuracy of a class was also defined as the ratio between the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard. However, F-measure can be computed as the harmonic mean of precision and recall, or 2 x Precision x Recall / (Precision + Recall).

\section*{Results}
Experimental evaluation of the classification of utterance sequences by using the proposed probabilistic models is reported with two set of experiments: the first set of experiment was focused on natural sequences, while the second set of experiment was performed on alternating sequences.

\subsection*{\textit{Predictive performance for natural PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}. It follows that first-order HMM demonstrates the best performance achieving the highest precision 0.7980 with 0.7989 F-measure. However, second-order Markov Chain demonstrates the lowest performance in terms of precision and F-measure. It can be seen that the lowest recall of 0.7686 is achieved by the first-order MC. On the other hand, second-order HMM achieves the highest recall 0.8449 and accuracy 0.8449. From Table~\ref{tab:result_norm_seq}, it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM. Precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. Overall, HMM performed better than MC for natural sequences. \\

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, first-order MC has the lowest performance compared to all other models across all metrics. Also, as in the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models. However, different from natural sequences, second-order MC demonstrates the highest precision 0.9778 as well as the highest F-measure of 0.9736. We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

@@

\subsection*{\textit{Impact of hidden states}}
Since the number of hidden states may impact the classification
performance, we use an optimum number of hidden states in HMM's. Figure~\ref{fig:hidden-states} shows the performance of HMMs with a different number of hidden states. The results illustrate that, in general, the state number didn't impact the classification performance significantly. All four models show their best performance when state number in between 4-8. Table~\ref{tab:emission_matrix} shows an example of the distribution of PPC codes over the optimum number of hidden states (4) for successful (S) and unsuccessful (U) HMMs. Log probability of each code reflects the state significance. In other words, these probabilities indicate that which hidden states are close to adolescent commitment language and which are not?

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
    \caption{\textbf{Performance of HMM's with different sequence type, order, and number of hidden states used in the experiment}.}
    \label{fig:hidden-states}
\end{figure}

\begin{table}[h]
\centering
\caption{Distribution of PPC codes over the optimal hidden states for successful (S) and unsuccessful (U) HMMs.}
\label{tab:emission_matrix}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Hidden} & \textbf{Log probability of each symbol (301, 117, 327, 343, 101, 305, 311, 331, 310, 328, 112,} \\ 
   \textbf{states} & \textbf{ 326, 302, 308, 314, 306, 119, 313, 333, 307, 325, 344, 109, 106, 332, 120, 345, 309, 312, 114,} \\
    & \textbf{ 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
S-3 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
S-4 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline

U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
U-3 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, \\
& -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, \\
& -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
U-4 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents show positive behavior change, the counselor immediately reflects on that positive tendency, to reinforce adolescent's desire to adhere to positive behavior change. On the other hand, general information by the counselor can lead to a negative behavior change in adolescents, even in the cases when adolescents were showing positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when presented with general information. \\

>>>>

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

\institutes{
$^1$Department of Computer Science,  
$^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics. In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose two methods, based on Markov chain and Hidden Markov Model, for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.  Our method achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences in predicting the outcome of motivational interviews with obese adolescents using first and second order Markov Chain and Hidden Markov Model, respectively. The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

@@

In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual annotation and examination of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our research addresses this problem from multiple angles. In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 1072 observed sequences, 880 were positive and 192 were negative. Alternating communication sequences are sub-sequences of a natural communication sequence, in which two consecutive utterances cannot belong to the same speaker (i.e. a counselor behavior code has to be followed by an adolescent behavior code and vice versa). For example, 6 alternating sequences (\{C1, A1, C2, A3\}, \{C1, A1, C3, A3\}, \{C1, A1, C4, A3\}, \{C1, A2, C2, A3\}, \{C1, A2, C3, A3\} and \{C1, A2, C4, A3\}) can be created from a natural sequence \{C1, A1, A2, C2, C3, C4, A3\}, where adolescent and counselor codes are represented by A and C, respectively. A total of 18753 alternating sequences have been observed in experimental dataset, out of which 18139 were successful and 614 were unsuccessful ones. For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences. Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
Natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
Alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
  \end{tabular}
\end{table}

@@

\subsection*{\textit{Prediction method}}

Generally, a sequence is a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as 117 $LUP+$ (low uptake, positive), 331 $OQ-ECT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are ``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.

Our proposed prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.

\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property). Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:

\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}

If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language. Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias. Figure~\ref{fig:delta} illustrates the accuracy of Markov Chain-based prediction method by varying the value of $\delta$. As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias. However, F-score sharply decreases after $\delta = -0.9$. Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.

@@

The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities. A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular state (symbol) $x_i$ is computed by looking at the k preceding states (symbols). Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols. In our experiment, we also used second order Markov model. Hence, our model has $N^2$ states with transition probability matrix of size $N^2 \times N$.

\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations. An HMM can be fully specified by the following quintuple: $\lambda = (N, M, A, B, \pi)$

@@

We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}. As in the case of MC, we also experimented with second-order HMM.

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, F-measure and accuracy using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). Accuracy of a class is defined as the ratio of the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.

\section*{Results}
Experimental evaluation of the proposed method is conducted on both natural sequences and alternating sequences.

\subsection*{\textit{Predictive performance for natural PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449) and accuracy (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

@@

\subsection*{\textit{Impact of hidden states}}
While MC is a non-parametric method, HMM has an important parameter, the number of hidden states. A natural question is how does this parameter affect the performance of HMM-based prediction  method. Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences. As follows from this Figure, in general, the number of states does not significantly impact the classification performance. All four models show their best performance when the number of states is between 4 and 8.

Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
    \label{fig:hidden-states}
\end{figure}

\begin{table}[!htb]
\centering
\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
\label{tab:emission_matrix_s}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
  \end{tabular}
\end{table}

\begin{table}[!htb]
\centering
\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
\label{tab:emission_matrix_u}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
  \end{tabular}
\end{table}

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation. Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews. \\
